### 2023-02-19 chapters/fairness/fif/conclusion.tex
connecting different chapters

In this chapter, we propose the Fairness Influence Function (FIF) to measure the effect of input features on the bias of classifiers on a given dataset. Our approach combines global sensitivity analysis and group-based fairness metrics in machine learning. Thereby, it is natural in our approach to formulate FIF of intersectional features, which together with individual FIFs explains bias with higher granularity. We theoretically analyze the properties of FIFs and provide an algorithm, {\fairXplainer}, for estimating FIFs using global variance decomposition and local regression. In experiments, {\fairXplainer} estimates individual and intersectional FIFs in real-world datasets and classifiers, approximates bias using FIFs with less estimation error than earlier methods, demonstrates a high correlation between FIFs and fairness interventions, and analyzes the impact of fairness enhancing and attack algorithms on FIFs. The results instantiate {\fairXplainer} as a global, granular, and more accurate explanation method to understand the sources of bias. Additionally, the resonance between the rules extracted by CORELS~\cite{rudin19stop} and the most influential features detected by {\fairXplainer} indicates that FIFs can be exploited in future to create an explainable proxy of a biased/unbiased classifier. Also, we aim to extend {\fairXplainer} to compute FIFs for complex data, such as image and text, and design algorithms leveraging FIFs to yield unbiased decisions.
%

>>>>

In this chapter, we propose the Fairness Influence Function (FIF) to measure the effect of input features on the bias of classifiers on a given dataset. Our approach combines global sensitivity analysis and group-based fairness metrics in machine learning. Thereby, it is natural in our approach to formulate FIF of intersectional features, which together with individual FIFs explains bias with higher granularity. We theoretically analyze the properties of FIFs and provide an algorithm, {\fairXplainer}, for estimating FIFs using global variance decomposition and local regression. In experiments, {\fairXplainer} estimates individual and intersectional FIFs in real-world datasets and classifiers, approximates bias using FIFs with less estimation error than earlier methods, demonstrates a high correlation between FIFs and fairness interventions, and analyzes the impact of fairness enhancing and attack algorithms on FIFs. The results instantiate {\fairXplainer} as a global, granular, and more accurate explanation method to understand the sources of bias. Additionally, the resonance between the rules extracted by CORELS~\cite{rudin2019stop} and the most influential features detected by {\fairXplainer} indicates that FIFs can be exploited in future to create an explainable proxy of a biased/unbiased classifier. Also, we aim to extend {\fairXplainer} to compute FIFs for complex data, such as image and text, and design algorithms leveraging FIFs to yield unbiased decisions.
%

### 2023-02-19 chapters/fairness/fif/experiments.tex
connecting different chapters

\subsection{Explainability and Applicability of FIFs}
\paragraph{Individual vs.\ Intersectional FIFs.} Now, we aim to understand the importance of intersectional FIFs over individual FIFs in Figure~\ref{fairness_fairXplainer_fig:fif_illustration_compas}. We consider COMPAS dataset with race $ \in $ \{Caucasian, non-Caucasian\} as the sensitive feature, and a logistic regression classifier to predict whether a person will re-offend crimes within the next two years. Since the classifier only optimizes training error, it demonstrates statistical parity of $ 0.174 $, i.e. it suggests that a non-Caucasian  has $ 0.174 $ higher probability of re-offending crimes than a Caucasian. Next, we investigate the sources of bias and present individual FIFs in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}, and both individual and intersectional FIFs in Figure~\ref{fairness_fairXplainer_fig:intersectional_fifs}. In both figures, we present influential features and their FIFs in the descending order of absolute values. According to {\fairXplainer}, the priors count (FIF $ = 0.1 $) dominates in increasing statistical parity\textemdash between Caucasian and non-Caucasian, their prior count demonstrates the maximum difference in the scaled variance of positive prediction. Other non-sensitive features have almost zero FIFs. However, in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}, higher-order FIFs ($ \lambda > 1 $) increases statistical parity by $ 0.031 $, denoting that the data is correlated and presenting only individual FIFs is not sufficient for understanding the sources of bias. For example, while both sex and age individually demonstrate almost zero influence on bias (Figure~\ref{fairness_fairXplainer_fig:individual_fifs}), their combined effect and intersectional effects with c-charge degree, priors count, and juvenile miscellaneous count contribute highly on statistical parity. In contrast to {\fairXplainer}, SHAP only estimates individual FIFs  (Figure~\ref{fairness_fairXplainer_fig:fif_shap}) and approximates statistical parity with higher error than {\fairXplainer}.  Interestingly, {\fairXplainer} yields FIF estimates of the classifier trained on COMPAS dataset that significantly matches with the rule-based classifier extracted from the COMPAS dataset using Certifiably Optimal Rule Lists (CORELS) algorithm~\cite[Figure 3]{rudin19stop}. From Figure 3 in~\cite{rudin19stop}, we observe that prior count is the only feature used individually to predict arrests, while age is paired with sex and prior counts respectively. While considering second-order intersectionality, {\fairXplainer} ($ \lambda = 2 $) yields priors count, (sex, age), and (age, priors count) as \emph{three of the top four features} that explains the observation of~\cite{rudin19stop} better than SHAP and {\fairXplainer} ($ \lambda = 1 $) considering only individual features. \emph{Therefore, {\fairXplainer} demonstrates a clearer understanding on the sources of bias of a classifier by simultaneously quantifying on intersectional influences and individual influences }.  %\todo{For a linear classifier with independent features, see if coefficient and FIFs are related.}

>>>>

\subsection{Explainability and Applicability of FIFs}
\paragraph{Individual vs.\ Intersectional FIFs.} Now, we aim to understand the importance of intersectional FIFs over individual FIFs in Figure~\ref{fairness_fairXplainer_fig:fif_illustration_compas}. We consider COMPAS dataset with race $ \in $ \{Caucasian, non-Caucasian\} as the sensitive feature, and a logistic regression classifier to predict whether a person will re-offend crimes within the next two years. Since the classifier only optimizes training error, it demonstrates statistical parity of $ 0.174 $, i.e. it suggests that a non-Caucasian  has $ 0.174 $ higher probability of re-offending crimes than a Caucasian. Next, we investigate the sources of bias and present individual FIFs in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}, and both individual and intersectional FIFs in Figure~\ref{fairness_fairXplainer_fig:intersectional_fifs}. In both figures, we present influential features and their FIFs in the descending order of absolute values. According to {\fairXplainer}, the priors count (FIF $ = 0.1 $) dominates in increasing statistical parity\textemdash between Caucasian and non-Caucasian, their prior count demonstrates the maximum difference in the scaled variance of positive prediction. Other non-sensitive features have almost zero FIFs. However, in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}, higher-order FIFs ($ \lambda > 1 $) increases statistical parity by $ 0.031 $, denoting that the data is correlated and presenting only individual FIFs is not sufficient for understanding the sources of bias. For example, while both sex and age individually demonstrate almost zero influence on bias (Figure~\ref{fairness_fairXplainer_fig:individual_fifs}), their combined effect and intersectional effects with c-charge degree, priors count, and juvenile miscellaneous count contribute highly on statistical parity. In contrast to {\fairXplainer}, SHAP only estimates individual FIFs  (Figure~\ref{fairness_fairXplainer_fig:fif_shap}) and approximates statistical parity with higher error than {\fairXplainer}.  Interestingly, {\fairXplainer} yields FIF estimates of the classifier trained on COMPAS dataset that significantly matches with the rule-based classifier extracted from the COMPAS dataset using Certifiably Optimal Rule Lists (CORELS) algorithm~\cite[Figure 3]{rudin2019stop}. From Figure 3 in~\cite{rudin2019stop}, we observe that prior count is the only feature used individually to predict arrests, while age is paired with sex and prior counts respectively. While considering second-order intersectionality, {\fairXplainer} ($ \lambda = 2 $) yields priors count, (sex, age), and (age, priors count) as \emph{three of the top four features} that explains the observation of~\cite{rudin2019stop} better than SHAP and {\fairXplainer} ($ \lambda = 1 $) considering only individual features. \emph{Therefore, {\fairXplainer} demonstrates a clearer understanding on the sources of bias of a classifier by simultaneously quantifying on intersectional influences and individual influences }.  %\todo{For a linear classifier with independent features, see if coefficient and FIFs are related.}

### 2023-02-19 chapters/fairness/fif/introduction.tex
connecting different chapters

The past few decades have witnessed a remarkable advancement in the field of machine learning, with its applications encompassing crucial decision-making processes such as college admissions~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring}, and more.  In such applications, the deployed machine learning classifier often demonstrates bias towards certain demographic groups present in the data~\cite{dwork2012fairness}. For instance, a machine learning classifier employed in college admissions may show a disproportionate preference for White-male candidates over Black-female candidates. This can occur due to historical biases present in the admission data, the classifier's accuracy-focused learning objective, or a combination of the two~\cite{landy1978correlates,zliobaite2015relation,berk2019accuracy}. Following such phenomena, multiple group-based fairness metrics, such as \textit{statistical parity}, \textit{equalized odds}, \textit{predictive parity}, etc.\ have been proposed to quantify the bias of the classifier on a dataset.  For example, a statistical parity of $ 0.6 $ for the college admission classifier would indicate that White-male candidates are offered admission $ 60\% $ more frequently than Black-female candidates~\cite{garg2020fairness,besse2021survey}.

Fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias. Motivated by the GDPR's ``right to explanation", research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}. In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset. To this end, we follow a global feature-attribution approach to explain the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier. In this context, existing bias attributing methods~\cite{begley2020explainability,lundberg2020explaining} are variants of local function approximation~\cite{sliwinski2019axiomatic}, whereas bias is a global statistical property of a classifier. Thus, \textit{we aim to design a bias attribution method that is global by construction}. In addition, existing methods only attribute the individual influence of features on bias while neglecting the \textit{intersectionality} among features. Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}. 
%\textit{By attributing bias to both individual and intersecting features, we can gain a more detailed understanding of bias} and pave the way for designing more effective fairness algorithms. 
In this chapter, \textit{we aim to design a global bias attribution framework and a corresponding algorithm that can quantify both the individual and intersectional influences of features leading to granular and functional explanation of the sources of bias.}

@@

\input{chapters/fairness/fif/figure_intro}
We illustrate the usefulness of our contributions via an example scenario proposed in\cite{ghosh2020justicia}.

@@

 %\red{This allows practitioners to deploy various affirmative or punitive actions.}\improvement{Examples?}
\begin{example}
	\label{fairness_fairXplainer_ex:motivating_example}  Following\cite{ghosh2020justicia}, we consider a classifier that decides an individual's eligibility for health insurance based on non-sensitive features: fitness  and income. Fitness and income depend on a sensitive feature age $ \in $ \{young, elderly\} leading to two sensitive groups, as highlighted in (Figure~\ref{fairness_fairXplainer_fig:dag_age_income_fitness}--\ref{fairness_fairXplainer_fig:distribution_example}).

>>>>

%The past few decades have witnessed a remarkable advancement in the field of machine learning, with its applications encompassing crucial decision-making processes such as college admissions~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring}, and more.  In such applications, the deployed machine learning classifier often demonstrates bias towards certain demographic groups present in the data~\cite{dwork2012fairness}. For instance, a machine learning classifier employed in college admissions may show a disproportionate preference for White-male candidates over Black-female candidates. This can occur due to historical biases present in the admission data, the classifier's accuracy-focused learning objective, or a combination of the two~\cite{landy1978correlates,zliobaite2015relation,berk2019accuracy}. Following such phenomena, multiple group-based fairness metrics, such as \textit{statistical parity}, \textit{equalized odds}, \textit{predictive parity}, etc.\ have been proposed to quantify the bias of the classifier on a dataset.  For example, a statistical parity of $ 0.6 $ for the college admission classifier would indicate that White-male candidates are offered admission $ 60\% $ more frequently than Black-female candidates~\cite{garg2020fairness,besse2021survey}.

In this chapter, we envision towards combining two research themes in the thesis: interpretability and fairness, and discuss an algorithmic framework to interpret or explain fairness in machine learning.
As demonstrated in Chapter~\ref{chapter:justicia} and~\ref{chapter:fvgm}, fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias. Motivated by the GDPR's ``right to explanation'', research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}. In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset. To this end, we follow a global feature-attribution approach to explain the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier. In this context, existing bias attributing methods~\cite{begley2020explainability,lundberg2020explaining} are variants of local function approximation~\cite{sliwinski2019axiomatic}, whereas bias is a global statistical property of a classifier. Thus, \textit{we aim to design a bias attribution method that is global by construction}. In addition, existing methods only attribute the individual influence of features on bias while neglecting the \textit{intersectionality} among features. Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}. 
%\textit{By attributing bias to both individual and intersecting features, we can gain a more detailed understanding of bias} and pave the way for designing more effective fairness algorithms. 
In this chapter, \textit{we aim to design a global bias attribution framework and a corresponding algorithm that can quantify both the individual and intersectional influences of features leading to granular and functional explanation of the sources of bias.}

@@

\input{chapters/fairness/fif/figure_intro}
We illustrate the usefulness of our contributions via an example scenario proposed in Example~\ref{fairness_justicia_example:intro}.

@@

 %\red{This allows practitioners to deploy various affirmative or punitive actions.}\improvement{Examples?}
\begin{example}
	\label{fairness_fairXplainer_ex:motivating_example} We consider a classifier that decides an individual's eligibility for health insurance based on non-sensitive features: fitness  and income. Fitness and income depend on a sensitive feature age $ \in $ \{young, elderly\} leading to two sensitive groups, as highlighted in (Figure~\ref{fairness_fairXplainer_fig:dag_age_income_fitness}--\ref{fairness_fairXplainer_fig:distribution_example}).

### 2023-02-19 chapters/fairness/fvgm/conclusion.tex
connecting different chapters

\section{Chapter Summary}
In this chapter, we propose {\fvgm}, an efficient fairness verification framework for linear classifiers based on a novel stochastic subset-sum problem. {\fvgm} encodes a graphical model of feature-correlations, represented as a Bayesian Network, and computes multiple group and causal fairness metrics accurately. We experimentally demonstrate that {\fvgm} is not only more accurate and scalable than the existing verifiers but also applicable in practical fairness tasks, such as verifying fairness attacks and enhancing algorithms, and computing the fairness influence functions. 
As a future work, we aim to design fairness-enhancing algorithms certified by fairness verifiers, such as {\fvgm}. %We also investigate a principled approach to detect the sources of bias by computing fairness influence functions (FIF). 
Since {\fvgm} serves as an accurate and scalable fairness verifier for linear classifiers, it will be interesting to design such verifiers for other ML models.
%Though the proposed solution with a Bayesian network has exponential worst-case complexity, it would be interesting to develop a pseudo-polynomial approximation in the future.

>>>>

\section{Chapter Summary}
In this chapter, we propose {\fvgm}, an efficient fairness verification framework for linear classifiers based on a novel stochastic subset-sum problem. {\fvgm} encodes a graphical model of feature-correlations, represented as a Bayesian Network, and computes multiple group and causal fairness metrics accurately. We experimentally demonstrate that {\fvgm} is not only more accurate and scalable than the existing verifiers but also applicable in practical fairness tasks, such as verifying fairness attacks and enhancing algorithms, and computing the fairness influence functions. In the next chapter, we elaborate on fairness influence functions as a framework to explaining group fairness metrics in machine learning.
\begin{comment}
	%As a future work, we aim to design fairness-enhancing algorithms certified by fairness verifiers, such as {\fvgm}. Since {\fvgm} serves as an accurate and scalable fairness verifier for linear classifiers, it will be interesting to design such verifiers for other machine learning models.
\end{comment}

### 2023-02-19 chapters/fairness/fvgm/introduction.tex
connecting different chapters

	Towards this goal, we particularly focus on the verification of \textit{linear classifiers} as our first step because there has been significant attention on linear classifiers in the context of fair algorithms \cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}. In this context, existing approaches suffer from limited scalability while verifying linear classifiers. This is due to the encoding of linear classifiers into SSAT/SMT formulas \cite{ghosh2020justicia,albarghouthi2017fairsquare}. For example, Justicia applies pseudo-Boolean to CNF translation of linear classifiers as a pre-processing step, and the encoding becomes large depending on the number of features and the precision of real-valued coefficients in linear classifiers.

>>>>

	Towards the goal of considering feature correlations in fairness verification, we particularly focus on the verification of \textit{linear classifiers} as our first step because of the significant attention on linear classifiers in fair algorithms \cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}. In this context, existing approaches suffer from limited scalability while verifying linear classifiers. This is due to the encoding of linear classifiers into SSAT/SMT formulas \cite{ghosh2020justicia,albarghouthi2017fairsquare}. For example, {\justicia} applies pseudo-Boolean to CNF translation of linear classifiers as a pre-processing step, and the encoding becomes large depending on the number of features and the precision of real-valued coefficients in linear classifiers.

### 2023-02-19 chapters/fairness/justicia/discussion.tex
highlevel pass
connecting different chapters

Our work opens up several new directions of research. One direction is to develop SSAT models and verifiers for popular classifiers like Deep networks and SVMs. Other direction is to develop SSAT solvers that can accommodate continuous variables and conditional probabilities by design.
%Of particular interest is the highlighted need for design of efficient SSAT solvers given that SSAT can be viewed as a natural target problem in the context of development of formal verification techniques for fairness.

>>>>

Our work opens up several new directions of research. One direction is to develop SSAT models and verifiers for popular classifiers like deep networks and SVMs. Other direction is to develop SSAT solvers that can accommodate continuous variables and conditional probabilities by design. Finally, an interesting direction is to enable {\justicia} to consider correlated feature distribution, which we pursue in the next chapter.
%Of particular interest is the highlighted need for design of efficient SSAT solvers given that SSAT can be viewed as a natural target problem in the context of development of formal verification techniques for fairness.

### 2023-02-19 chapters/fairness/justicia/practical.tex
connecting different chapters

\subsubsection{Extension to Continuous Features}
In practical problems, features are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM2020}, are generally trained on a Boolean abstraction of input features. In order to perform the Boolean abstraction, each categorical feature is one-hot encoded and each real-valued feature is discretized into a set of Boolean features~\cite{LKCL2019,GMM2020}.

>>>>

\subsubsection{Extension to Continuous Features}
In practical problems, features are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM2020}, are generally trained on a Boolean abstraction of input features. In order to perform the Boolean abstraction, each categorical feature is one-hot encoded and each real-valued feature is discretized into a set of Boolean features~\cite{lakkaraju2019faithful,GMM2020}.

### 2023-02-19 chapters/interpretability/imli/appendix.tex
connecting different chapters

In Figure~\ref{fig:interpretability_imli_additional_fig:all_preformances_non_incremental_vs_incremental},

>>>>

In Figure~\ref{fig:interpretability_imli_additional_fig:all_preformances_non_incremental_vs_incremental}, we compare both non-incremental and incremental encoding in terms of median training time, rule-size, test accuracy, and train accuracy over ten cross validation folds. In all datasets, the incremental encoding demonstrates higher efficiency than the non-incremental encoding by requiring less training time. In large datasets, such as Magic, Credit, and Adult, the non-incremental encoding always times out, thereby producing zero-size rules because of failing to solve the MaxSAT query. As a result, the accuracy of the non-incremental encoding is also less than the incremental encoding in large datasets. In small datasets, the incremental encoding demonstrates a competitive accuracy and rule-size with the non-incremental encoding. Therefore, the incremental encoding in {\imli} demonstrates its effectiveness in training time, accuracy, and rule-size over the non-incremental encoding.

### 2023-02-19 chapters/interpretability/imli/conclusion.tex
all papers added
minor edits in many parts of the thesis
connecting different chapters

\section{Conclusion}
\label{interpretability_imli_sec:conclusion}
Interpretable machine learning is gaining more focus with applications in many safety-critical domains. Considering the growing demand for interpretable models, it is challenging to design learning frameworks that satisfy all aspects: being accurate, interpretable, and scalable in practical classification tasks.  In this paper, we have proposed a MaxSAT-based framework {\imli} for learning interpretable rule-based classifiers expressible in CNF formulas. {\imli} is built on efficient integration of incremental learning, specifically mini-batch and iterative learning, with MaxSAT-based formulation.  In our empirical evaluation, {\imli} achieves the best balance among prediction accuracy, interpretability, and scalability. In particular, {\imli} demonstrates competitive prediction accuracy and rule-size compared to existing interpretable rule-based classifiers. In addition, {\imli} achieves impressive scalability than both interpretable and non-interpretable classifiers by learning interpretable rules on million-size datasets with higher accuracy.  Finally, {\imli}  generates other popular interpretable classifiers such as decision lists and decision sets using the same framework.

>>>>

\section{Chapter Summary}
\label{interpretability_imli_sec:conclusion}
Interpretable machine learning is gaining more focus with applications in many safety-critical domains. Considering the growing demand for interpretable models, it is challenging to design learning frameworks that satisfy all aspects: being accurate, interpretable, and scalable in practical classification tasks.  In this chapter, we have proposed a MaxSAT-based framework {\imli} for learning interpretable rule-based classifiers expressible in CNF formulas. {\imli} is built on efficient integration of incremental learning, specifically mini-batch and iterative learning, with MaxSAT-based formulation.  In our empirical evaluation, {\imli} achieves the best balance among prediction accuracy, interpretability, and scalability. In particular, {\imli} demonstrates competitive prediction accuracy and rule-size compared to existing interpretable rule-based classifiers. In addition, {\imli} achieves impressive scalability than both interpretable and non-interpretable classifiers by learning interpretable rules on million-size datasets with higher accuracy.  Finally, {\imli}  generates other popular interpretable classifiers such as decision lists and decision sets using the same framework. An interesting direction is to leverage the incremental learning in {\imli} to learn more expressible yet interpretable classification rules, which we explore in the next chapter.

### 2023-02-19 chapters/interpretability/relaxed-cnf/ilp_formulation.tex
connecting different chapters

\textbf{Relaxing the ILP problem:}
\label{interpretability_crr_sec:relaxation}
The ILP query $ Q $ has binary integer constraints and the solution of this integer program is computationally intractable. 
A common approach that efficiently finds an approximate solution to such a problem extends to relax the integer constraints,  solves the LP-relaxed (linear programming relaxation) problem, and then rounds the non-integer variables to get an integer solution as in~\cite{MV2013}. 
In our case, we cannot relax all integer constraints because it may violate the structure of relaxed-CNF rules. Specifically, $ \eta_c $ and $ \eta_l $ (or $ \eta_{l,i} $) must be integers in the construction of relaxed-CNF rules, and $ \noise_{q,i} $ needs to be an integer to precisely calculate the noise variable $ \noise_q $.  However,  we can relax feature variable $ \bool^i_j $ and noise variable $ \noise_q $ and solve the corresponding MILP  problem.  To construct the MILP problem, we replace Eq.~\ref{interpretability_crr_eq:ilp_range} with $ 0 \le \bool^i_j \le 1 $ and Eq.~\ref{interpretability_crr_eq:ilp_range_2} with $ 0 \le \noise_q \le 1 $, and the rest of the constraints in $ Q $ remain the same.  If non-zero $ \bool^i_j $ is found in the solution, we set it to $ 1 $ and then construct the rule according to Construction~\ref{interpretability_crr_cons:rule}.

@@

After applying entropy-based discretization on continuous features,  the dataset  contains only categorical features, that can be converted to binary features using one-hot encoding as in \cite{LKCL2019}. In this encoding, a Boolean vector is introduced with cardinality equal to the number of distinct categories. Let a categorical feature have three  categories `red',`green', and `yellow'. In one-hot encoding, samples with category-value `red',`green', and `yellow' would be converted into binary features while taking values $ 100 $, $ 010 $, and $ 001 $, respectively.

>>>>

\textbf{Relaxing the ILP problem:}
\label{interpretability_crr_sec:relaxation}
The ILP query $ Q $ has binary integer constraints and the solution of this integer program is computationally intractable. 
A common approach that efficiently finds an approximate solution to such a problem extends to relax the integer constraints,  solves the LP-relaxed (linear programming relaxation) problem, and then rounds the non-integer variables to get an integer solution as in~\cite{malioutov2013exact}. 
In our case, we cannot relax all integer constraints because it may violate the structure of relaxed-CNF rules. Specifically, $ \eta_c $ and $ \eta_l $ (or $ \eta_{l,i} $) must be integers in the construction of relaxed-CNF rules, and $ \noise_{q,i} $ needs to be an integer to precisely calculate the noise variable $ \noise_q $.  However,  we can relax feature variable $ \bool^i_j $ and noise variable $ \noise_q $ and solve the corresponding MILP  problem.  To construct the MILP problem, we replace Eq.~\ref{interpretability_crr_eq:ilp_range} with $ 0 \le \bool^i_j \le 1 $ and Eq.~\ref{interpretability_crr_eq:ilp_range_2} with $ 0 \le \noise_q \le 1 $, and the rest of the constraints in $ Q $ remain the same.  If non-zero $ \bool^i_j $ is found in the solution, we set it to $ 1 $ and then construct the rule according to Construction~\ref{interpretability_crr_cons:rule}.

@@

After applying entropy-based discretization on continuous features,  the dataset  contains only categorical features, that can be converted to binary features using one-hot encoding as in \cite{lakkaraju2019faithful}. In this encoding, a Boolean vector is introduced with cardinality equal to the number of distinct categories. Let a categorical feature have three  categories `red',`green', and `yellow'. In one-hot encoding, samples with category-value `red',`green', and `yellow' would be converted into binary features while taking values $ 100 $, $ 010 $, and $ 001 $, respectively.

### 2023-02-19 chapters/introduction.tex
connecting different chapters

\subsection*{Explaining Fairness Metrics: Identifying Sources of Bias}
Fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias. Motivated by the GDPR's ``right to explanation", research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}. In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset. To this end, we follow a global feature-attribution approach to explain the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier. In this context, existing bias attributing methods~\cite{begley2020explainability,lundberg2020explaining} are variants of local function approximation~\cite{sliwinski2019axiomatic}, whereas bias is a global statistical property of a classifier. Thus, \textit{we aim to design a bias attribution method that is global by construction}. In addition, existing methods only attribute the individual influence of features on bias while neglecting the \textit{intersectionality} among features. Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}.  In our contribution, we propose an algorithm, called {\fairXplainer}, based on global sensitivity analysis to estimate the fairness influence function (FIF) of individual and intersectional features. In experiments, {\fairXplainer} approximates bias based on FIFs more accurately than existing methods. We also demonstrate higher correlation of FIFs with fairness interventions, thereby demonstrating the importance of FIFs in designing improved fairness algorithms in future.

>>>>

\subsection*{Explaining Fairness Metrics: Identifying Sources of Bias}
Fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias. Motivated by the GDPR's ``right to explanation'', research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}. In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset. To this end, we follow a global feature-attribution approach to explain the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier. In this context, existing bias attributing methods~\cite{begley2020explainability,lundberg2020explaining} are variants of local function approximation~\cite{sliwinski2019axiomatic}, whereas bias is a global statistical property of a classifier. Thus, \textit{we aim to design a bias attribution method that is global by construction}. In addition, existing methods only attribute the individual influence of features on bias while neglecting the \textit{intersectionality} among features. Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}.  In our contribution, we propose an algorithm, called {\fairXplainer}, based on global sensitivity analysis to estimate the fairness influence function (FIF) of individual and intersectional features. In experiments, {\fairXplainer} approximates bias based on FIFs more accurately than existing methods. We also demonstrate higher correlation of FIFs with fairness interventions, thereby demonstrating the importance of FIFs in designing improved fairness algorithms in future.
